---
title: "INF 6490 Final Project"
author: "Fred LaBeau"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries for assessment
```{r}
library(tidyverse)  
library(janitor)    
library(skimr)      
library(GGally)     
library(broom)      
library(lubridate)  
library(car)        
library(rstatix)    
library(effectsize) 
library(emmeans)  
library(rsample)   
library(yardstick)

```

#Import data and look at it
```{r}
ai_jobs_raw <- readxl::read_xlsx("ai_job_market.xlsx")

glimpse(ai_jobs_raw)
head(ai_jobs_raw)
```


#Clean job data
```{r}
ai_jobs <- ai_jobs_raw %>%
  janitor::clean_names()
```


#Parse salary range and divide into 3 sections (salary_min, salary_max, salary_mid)
```{r}
ai_jobs <- ai_jobs %>%
  tidyr::separate(
    col   = salary_range_usd,
    into  = c("salary_min", "salary_max"),
    sep   = "-",
    remove = TRUE
  ) %>%
  mutate(
    salary_min = as.numeric(salary_min),
    salary_max = as.numeric(salary_max),
    salary_mid = (salary_min + salary_max) / 2
  )
```


#Convert date and categorical variables
```{r}
ai_jobs <- ai_jobs %>%
  mutate(
    #Convert posted_date to Date
    posted_date = as.Date(posted_date),

    #Create ordered factor for experience level
    experience_level = factor(
      experience_level,
      levels = c("Entry", "Mid", "Senior"),
      ordered = TRUE
    ),

    #Set employment type as factor
    employment_type = factor(employment_type),
    
    #Set Industry as factor
    industry = factor(industry),

    #Set company size as ordered factor (dataset uses: Startup, Mid, Large)
    company_size = factor(
      company_size,
      levels = c("Startup", "Mid", "Large"),
      ordered = TRUE
    ),

    #Set job title as factor
    job_title = factor(job_title)
  )
```


#Additional dataset checking
```{r}
#Overview of variables and missing values
skimr::skim(ai_jobs)

#Count NAs per column to clean
colSums(is.na(ai_jobs))


glimpse(ai_jobs)
```


#Compute measures of central tendency and dispersion
```{r}
#Identify continuous (numeric) variables ----
numeric_vars <- ai_jobs %>%
  dplyr::select(where(is.numeric))

glimpse(numeric_vars)

#Compute the statistical mode ----
get_mode <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

#Summary: mean, median, variance, SD for each numeric variable ----
numeric_summary <- numeric_vars %>%
  summarise(
    across(
      everything(),
      list(
        mean   = ~ mean(. , na.rm = TRUE),
        median = ~ median(. , na.rm = TRUE),
        var    = ~ var(. , na.rm = TRUE),
        sd     = ~ sd(. , na.rm = TRUE)
      )
    )
  ) %>%
  #Make the output tidy: one row per variable/stat combination
  pivot_longer(
    cols      = everything(),
    names_to  = c("variable", "stat"),
    names_sep = "_",
    values_to = "value"
  ) %>%
  arrange(variable, stat)

numeric_summary

#Add mode for each numeric variable ----
numeric_mode <- numeric_vars %>%
  summarise(
    across(
      everything(),
      get_mode,
      .names = "{.col}_mode"
    )
  ) %>%
  pivot_longer(
    cols      = everything(),
    names_to  = "variable",
    values_to = "mode"
  ) %>%
  mutate(
    variable = stringr::str_remove(variable, "_mode$")
  )

numeric_mode

```


#Generate frequency tables and proportions for variables
```{r}
#Choose categorical variables
categorical_vars <- ai_jobs %>%
  dplyr::select(experience_level, industry, employment_type, company_size)

glimpse(categorical_vars)

#Create frequency table + proportions based on experience level
experience_table <- ai_jobs %>%
  janitor::tabyl(experience_level) %>%
  janitor::adorn_totals("row") %>%
  janitor::adorn_percentages("col") %>%
  janitor::adorn_pct_formatting(digits = 1)

experience_table

#Create frequency table + proportions based on industry
industry_table <- ai_jobs %>%
  janitor::tabyl(industry) %>%
  janitor::adorn_totals("row") %>%
  janitor::adorn_percentages("col") %>%
  janitor::adorn_pct_formatting(digits = 1)

industry_table

#Create frequency table + proportions based on employment type
employment_type_table <- ai_jobs %>%
  janitor::tabyl(employment_type) %>%
  janitor::adorn_totals("row") %>%
  janitor::adorn_percentages("col") %>%
  janitor::adorn_pct_formatting(digits = 1)

employment_type_table

#Create frequency table + proportions based on company size
company_size_table <- ai_jobs %>%
  janitor::tabyl(company_size) %>%
  janitor::adorn_totals("row") %>%
  janitor::adorn_percentages("col") %>%
  janitor::adorn_pct_formatting(digits = 1)

company_size_table

```

#Histogram creation to show salary distribution

```{r}
ggplot(ai_jobs, aes(x = salary_mid)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Median AI Job Salaries",
    x = "Median Salary (USD)",
    y = "Frequency"
  )

```

#Bar chart creation to show job experience levels
```{r}
ggplot(ai_jobs, aes(x = experience_level)) +
  geom_bar(fill = "darkorchid") +
  labs(
    title = "Count of AI Jobs by Experience Level",
    x = "Experience Level",
    y = "Number of Job Listings"
  )

```

#Box plot for salary by experience level
```{r}
ggplot(ai_jobs, aes(x = experience_level, y = salary_mid)) +
  geom_boxplot(fill = "tan") +
  labs(
    title = "Salary Distribution by Experience Level",
    x = "Experience Level",
    y = "Median Salary (USD)"
  )

```

#Box plot of salary by industry
```{r}
ggplot(ai_jobs, aes(x = industry, y = salary_mid)) +
  geom_boxplot(fill = "mediumpurple") +
  labs(
    title = "Salary Distribution by Industry",
    x = "Industry",
    y = "Median Salary (USD)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#Box plot of Salary by employment type
```{r}
ggplot(ai_jobs, aes(x = employment_type, y = salary_mid)) +
  geom_boxplot(fill = "seagreen") +
  labs(
    title = "Salary Distribution by Employment Type",
    x = "Employment Type",
    y = "Median Salary (USD)"
  )

```

#Box plot of Salary vs Company Size
```{r}
ggplot(ai_jobs, aes(x = company_size, y = salary_mid)) +
  geom_boxplot(fill = "tomato") +
  labs(
    title = "Salary Distribution by Company Size",
    x = "Company Size",
    y = "Median Salary (USD)"
  )

```

#Correlation Visualization
```{r}
ggplot(ai_jobs, aes(x = experience_level, y = salary_mid)) +
  geom_boxplot(fill = "lightblue") +
  facet_wrap(~ industry, scales = "free_y") +
  labs(
    title = "Salary by Experience Within Each Industry",
    x = "Experience Level",
    y = "Median Salary (USD)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#Checking assumptions for salary comparisons
```{r}

ai_jobs %>%
  summarise(
    n = sum(!is.na(salary_mid)),
    mean = mean(salary_mid, na.rm = TRUE),
    sd = sd(salary_mid, na.rm = TRUE)
  )


ai_jobs %>%
  group_by(experience_level) %>%
  summarise(p_shapiro = shapiro.test(salary_mid)$p.value)

# Homogeneity of variance Levene’s Test1
car::leveneTest(salary_mid ~ experience_level, data = ai_jobs)

# Homogeneity of variance for industry
car::leveneTest(salary_mid ~ industry, data = ai_jobs)

# Homogeneity of variance for employment type
car::leveneTest(salary_mid ~ employment_type, data = ai_jobs)

```

#One way ANOVA
```{r}
# ANOVA
anova_rq1 <- aov(salary_mid ~ experience_level, data = ai_jobs)
summary(anova_rq1)

# Effect size (eta squared) with CI
effectsize::eta_squared(anova_rq1, ci = 0.95)

```

#Estimate marginal means
```{r}
emm_rq1 <- emmeans::emmeans(anova_rq1, ~ experience_level)
emm_rq1
confint(emm_rq1, level = 0.95)
pairs(emm_rq1, adjust = "tukey")

```

#Assessing whether industries pay differently
```{r}
anova_rq3 <- aov(salary_mid ~ industry * experience_level, data = ai_jobs)
summary(anova_rq3)

# Effect sizes with CI
effectsize::eta_squared(anova_rq3, ci = 0.95)

emm_industry <- emmeans::emmeans(anova_rq3, ~ industry)
pairs(emm_industry, adjust = "tukey")
confint(emm_industry, level = 0.95)

emm_ind_by_exp <- emmeans::emmeans(anova_rq3, ~ industry | experience_level)
pairs(emm_ind_by_exp, adjust = "tukey")
confint(emm_ind_by_exp, level = 0.95)


```

#Assessing if employment type affects salary
```{r}
anova_rq4 <- aov(salary_mid ~ employment_type, data = ai_jobs)
summary(anova_rq4)

effectsize::eta_squared(anova_rq4, ci = 0.95)

emm_emp <- emmeans::emmeans(anova_rq4, ~ employment_type)
confint(emm_emp, level = 0.95)
pairs(emm_emp, adjust = "tukey")


```

#Chi square to assess relationship of categories
```{r}
tab_emp_ind <- table(ai_jobs$employment_type, ai_jobs$industry)
chisq.test(tab_emp_ind)
effectsize::cramers_v(tab_emp_ind, ci = 0.95)

```

#Spearman correlation between salary and experience level
```{r}
ai_jobs <- ai_jobs %>%
  mutate(exp_rank = as.numeric(experience_level))

cor.test(ai_jobs$salary_mid, ai_jobs$exp_rank, method = "spearman")

```

```{r}
set.seed(123)


model_df <- ai_jobs %>%
  select(
    salary_mid, experience_level, industry, employment_type, company_size, job_title
  ) %>%
  drop_na()


split_obj <- initial_split(model_df, prop = 0.80, strata = experience_level)
train_df  <- training(split_obj)
test_df   <- testing(split_obj)

nrow(train_df)
nrow(test_df)

```

#Linear regression
```{r}
lin_mod <- lm(
  salary_mid ~ experience_level + industry + employment_type + company_size + job_title,
  data = train_df
)

summary(lin_mod)
broom::tidy(lin_mod)
broom::glance(lin_mod)

test_pred <- test_df %>%
  mutate(
    pred_salary_mid = predict(lin_mod, newdata = test_df)
  )

head(test_pred)

lin_metrics <- test_pred %>%
  yardstick::metrics(truth = salary_mid, estimate = pred_salary_mid) %>%
  filter(.metric %in% c("rmse", "rsq"))

lin_metrics

par(mfrow = c(2,2))
plot(lin_mod)
par(mfrow = c(1,1))


```

